---
title: "Fireplace Code"
author: "Allison O'Connor"
date: "December 12, 2014"
output: html_document
---
This file pulls together the various code files used in the Fireplaces: Heart of the Home project.

The first tests for converting the data used a short file in order to see what issues came up, then used progressively larger files to find other issues.

There were two separate phases: getting the layout file, which describes the fields and their length, into usable shape, and then using that information to parse the data file into fields. Not all entries in the layout file included information about the length of the field, so the first step was to filter out the layout entries that do not have a value in the LENGTH field.

```{r}
layout <- read.csv("PDLayout.csv", stringsAsFactors = FALSE)
library(dplyr)
layout <- layout %>%
filter(!is.na (LENGTH))

```

Then, read the fixed width format data into a file. This is a test file to start with. In one of the tests, the presence of # and tabs in the data required the above mentioned manual clean up. This is where in the process I discovered that issue.

```{r}
PDData <- read.fwf("firstrowsB.txt", widths = layout$LENGTH, col.names = layout$COLUMN.NAME)
head(PDData)
```

Here all the chunked files get read in, then the separate pieces get bound together to work as one.

```{r}
PDDataA <- read.fwf("cleanedFFXDataAA.txt", widths = layout$LENGTH, col.names = layout$COLUMN.NAME,
                    stringsAsFactors = FALSE)
PDDataB <- read.fwf("cleanedFFXDataAB.txt", widths = layout$LENGTH, col.names = layout$COLUMN.NAME,  stringsAsFactors = FALSE)
PDDataC <- read.fwf("cleanedFFXDataAC.txt", widths = layout$LENGTH, col.names = layout$COLUMN.NAME,  stringsAsFactors = FALSE)
PDDataD <- read.fwf("cleanedFFXDataAD.txt", widths = layout$LENGTH, col.names = layout$COLUMN.NAME,  stringsAsFactors = FALSE)
PDDataE <- read.fwf("cleanedFFXDataAE.txt", widths = layout$LENGTH, col.names = layout$COLUMN.NAME,  stringsAsFactors = FALSE)
PDDataF <- read.fwf("cleanedFFXDataAF.txt", widths = layout$LENGTH, col.names = layout$COLUMN.NAME,  stringsAsFactors = FALSE)
PDDataG <- read.fwf("cleanedFFXDataAG.txt", widths = layout$LENGTH, col.names = layout$COLUMN.NAME,  stringsAsFactors = FALSE)
PDDataH <- read.fwf("cleanedFFXDataAH.txt", widths = layout$LENGTH, col.names = layout$COLUMN.NAME,  stringsAsFactors = FALSE)

# take all the data pieces and bind them into one object, then remove all the little objects to free up memory.

FFXdata <- rbind(PDDataA, PDDataB, PDDataC, PDDataD, PDDataE, PDDataF, PDDataG, PDDataH)
rm(PDDataA, PDDataB, PDDataC, PDDataD, PDDataE, PDDataF, PDDataG, PDDataH)
```

With all the data loaded into R, it is possible to perform more cleanup to the data. The county data includes all parcels, and has 360,038 observations of 142 variables. We need to remove any non-residential records. We also want to remove records with data anomolies such as a house built in a year that has not yet occurred, select only the fields we are most interested in, and give them easier to understand field names. We can also use the mutate function to change character data into integers, and create a column that will give us the decade a house was built. We end up with 282,551 observations of 15 variables.

```{r}
# these are the codes for various residential land uses, which I will need further down. 
residential_codes <- c(011, 012, 013, 014, 015, 021, 029, 031, 032, 033, 034, 035, 036, 037, 039, 040, 041, 042, 043, 044, 045, 046, 047, 048, 049)

FFX_res_data <- FFXdata %>%
 filter(LUC%in% residential_codes) %>%
 select(parcel_id = PARID,
        address_number = ADRNO,
        address_number_suffix = ADRADD,
        street_direction = ADRDIR,
        street_name = ADRSTR,
        street_suffix = ADRSUF,
        city_name = CITYNAME,
        zip = ZIP1,
        sale_price = PRICE,
        sale_date = SALEDT,
        fireplaces = WBFP_O...WBFP_PF,
        year_built = YRBLT,
        land_use_code = LUC,
        sqft_living_area = SFLA) %>%
  mutate(decade = round_year(year_built, round_to = "decade")) %>%
  mutate(fireplaces = as.integer(fireplaces)) %>%
  mutate(sqft_living_area = as.integer(sqft_living_area)) %>%
  filter(year_built < 2100) 
  tbl_df(FFX_res_data)
```


```{r}
#This forces the sqft_living_area to be an integer. For some reason the mutate above does not force it to be an integer
FFX_res_data <- FFX_res_data %>%
  mutate(sqft_living_area = as.integer(sqft_living_area))
```
```{r}
load("../FFXData/FFXDAta.RData")
library(dplyr)
library(ggplot2)
#keep ggplot in here for plots in files where this code is included.

#size of house, pre 1840
house_size <- FFX_res_data %>%
  select(year_built, sqft_living_area, decade) %>%
  filter(year_built < 1840)

#x number fireplaces and size of house
fp_sqft <- FFX_res_data %>%
  select(fireplaces, year_built, sqft_living_area, decade)%>%
  filter(year_built < 1840)



#calculate how many houses have x number of fireplaces
FFX_number_fp_per_house <-FFX_res_data %>%
  select(fireplaces, decade) %>%
  group_by(decade, fireplaces) %>%
  summarize(houses_with_x_fps = n())

#calculate how many houses were built per decade
FFX_houses_built_per_decade <- FFX_res_data %>%
  group_by(decade) %>%
  summarize(houses_built = n())

#join two tables together
FFX_number_fp_per_house <- FFX_number_fp_per_house %>%
  left_join(FFX_houses_built_per_decade, by = "decade")

#calculate percentage
FFX_number_fp_per_house <- FFX_number_fp_per_house %>%
  mutate(percentage = houses_with_x_fps/houses_built)
  
```
  
 