---
title: "Fireplaces: Heart of the Home"
author: "Allison O'Connor"
date: "December 12, 2014"
output: html_document
---

##Summary

Fireplaces are the proverbial heart of the home. They served critical functions in early American homes, but were plauged with difficulties, inefficiencies, and safety concerns. These issues spurred inventions and innovations which replaced the fireplace in performing the critical functions it once served. Yet, even though other technologies replaced the functions of the fireplace, the fireplace remained a feature in American homes. This study examines tax records from Fairfax County, VA, and 1940 national census records regarding central heating, to better understand fireplaces in American homes.

##Introduction

In the small spaces of typical early American homes, one hearth served many functions, fulfilling physical needs related to warmth, light, and food preparation, as well as spiritual and emotional needs related to companionship, fellowship, entertainment, and the satisfaction that comes from being close to a domestic fire. The hearth represented a place where one might find comfort and solace, even as it was the location of hard physical labor.

It was also the source of much frustration and consternation: smoky chimneys, inefficient use of fuel, an inability to warm a room evenly, the risks to life and property from flying sparks, and the hard physical labor involved in cooking and heating with a fireplace. Over time, numerous inventors sought to alleviate one or more of these problems and make life better through home heating technologies—including modifications to fireplaces, sheet or cast iron stoves, and furnaces—and cooking technologies—including the cookstove and range. Even as technological solutions made the fireplace unnecessary to maintain quality of life, the importance of the heritage of the hearth did not diminish.

There were a number of significant issues that made the hearth as much a source of concern as it was of comfort. Wood fuel became scarce and expensive. Furthermore, the supply could be unreliable—particularly in areas of established settlement and denser population or during times of extreme winter weather.^[Sean Patrick Adams, “Warming the Poor and Growing Consumers: Fuel Philanthropy in the Early Republic’s Urban North,” *Journal of American History* 95, no. 1 (June 2008): 69–94; that wood fuel was expensive and in short supply in areas outside the urban North, see Charles E. Peterson, “Early House-Warming by Coal-Fires,” *Journal of the Society of Architectural Historians* 9, no. 4 (December 1, 1950): 21–24, doi:10.2307/987472.]  Even so, Americans were slow to convert to the “close” stove fires that saved fuel. In 1792, while seeking to answer the question of “whether *close* or *open* fire-places are the wholesomest,” one writer noted that, “Prejudice among the Americans is great against those that are *close*, such as stoves.”^[“Observations on the Effects of Close and Open Fire Places,” *American Museum, Or, Universal Magazine* 12, no. 6 (December 1792): 351–52.]  Anglo-Americans preferred the open fireplaces of their British ancestors. The Pennsylvania Germans, by contrast, were extensive users of stoves.^[Ibid.] 

Safety was always a concern. Individuals needed to pay close attention to the hearth, lest any stray sparks or embers ignited an uncontrolled fire that could easily spread, causing significant property damage and possibly loss of life. In October 1711, a fire broke out in Boston, burning down the “town-house … the old meeting house … near about an hundred” houses and an unspecified number of shops; remains of seven or eight people who died in the fire were recovered.^[*Advice from Taberah. A Sermon Preached after the Terrible Fire, Which, (attended with Some Very Lamentable and Memorable Circumstances, on Oct. 2,3. 1711.) Laid a Considerable Part of Boston, in Ashes. Directing a Pious Improvement of Every Calamity, but More Especially of so Calamitous a Desolation. By Cotton Mather, D.D. [Two Lines from Numbers]*. (Boston in N.E., 1711), http://find.galegroup.com/ecco/infomark.do?&source=gale&prodId=ECCO&userGroupName=viva_gmu&tabID=T001&docId=CB3326167784&type=multipage&contentSet=ECCOArticles&version=1.0.]     The fire on December 9, 1747, caused Boston officials to set aside a day of fasting and prayer.^[*The Cry of Oppression Where Judgment Is Looked For, and the Sore Calamities Such a People May Expect from a Righteous God: Illustrated in Two Discourses from Isaiah V. Vii. on January 28. 1747,48: Which Was Set Apart by the Government for Fasting and Prayer, in Consideration of the Remarkable Judgments of God upon the Land: And More Especially the Destruction of the Court-House by Fire the 9th of December Last. By Nathaniel Appleton, A.M. Pastor of the First Church in Cambridge. Publish’d at the Desire of a Great Number of the Hearers. [Seven Lines from Isaiah]*. (Boston: New-England, 1748), http://find.galegroup.com/ecco/infomark.do?&source=gale&prodId=ECCO&userGroupName=viva_gmu&tabID=T001&docId=CB3329252243&type=multipage&contentSet=ECCOArticles&version=1.0.]   A fire in March 1760, again in Boston, destroyed 174 dwellings, a similar number of warehouses, shops and other buildings, and prompted the Massachusetts governor to send out a call for charity and relief from those who could provide it to the 220 families in need.^[Massachusetts.Governor (1757-1760 : Pownall), *By His Excellency Thomas Pownall, Esq; ... A Brief. It Having Pleased Almighty God to Permit a Fire to Break out in the Town of Boston, on the 20th Instant ... Given at the Council-Chamber in Boston the Twenty-Fourth Day of March, 1760 ...* (Boston, 1760), http://find.galegroup.com/ecco/infomark.do?&source=gale&prodId=ECCO&userGroupName=viva_gmu&tabID=T001&docId=CB3330923339&type=multipage&contentSet=ECCOArticles&version=1.0.]  By May 6, 1760, the governor of Maryland forwarded the call for charity to aid those who suffered losses.^[  Maryland.Governor (1753-1769 : Sharpe), *By His Excellency Horatio Sharpe, Esq, Governor and Commander in Chief in and over the Province of Maryland. A Brief. It Having Been Represented to Me, by His Majesty’s Governor of the Massachusetts-Bay in New-England, That on the 20th of March Last, a Fire Broke out in Th Town of Boston, in Such Manner as to Elude All Means for Suppressing the Same, until It Had (according to the Best Information That Could Be Obtained) Destroyed 174 Dwelling-Houses, and as Many Warehouses, Shops and Other Buildings ...* ([Annapolis ], 1760), http://find.galegroup.com/ecco/infomark.do?&source=gale&prodId=ECCO&userGroupName=viva_gmu&tabID=T001&docId=CB3330923324&type=multipage&contentSet=ECCOArticles&version=1.0.]  A fire in Boston in July 1794 lead to the destruction of “upwards of 45 houses, stores and shops,” in approximately three hours, leaving “upwards of 100 families” homeless.^[“Boston, July 31. Fire! Fire!,” *United States Magazine, Or, General Repository of Useful Instruction & Rational Amusement* 1 (July 1794): 254–254.]  Editors of newspapers and other periodicals printed directions for reducing the likelihood of “calamities by fire.”^[“Directions for Preventing Calamities by FIRE. Recommended by the ‘Massachusetts Fire Society;’ with Notes and Additional Directions by the Editor.,” *Archives of Useful Knowledge* 3, no. 2 (October 1812): 159–63; For a review of a book or pamphlet on the subject, see “Art. 31. Various Methods to Prevent Fires in Houses and Shipping, and for Preserving the Lives of People at Fires,” *Monthly Review* 53 (September 1775): 266–266.]  Extrapolating from the records for Boston to apply the frequency of catastrophic fires to other cities throughout the colonies and early republic, fire was a significant risk. Any inventions that helped mitigate those risks by assisting in the containment or control of domestic fires likely found a receptive audience. Benjamin Franklin, in his An Account of the New Invented Pennsylvanian Fire-places, considered the ability to do both of these as great advantages of his Pennsylvanian fireplace over regular fireplaces. 

Books written for children in the early nineteenth century emphasized the dire consequences that might come if a child failed to heed the cautions of his or her elders. The story of *Little Jane* told the tale of a girl six-years old, who, unable to resist the temptation to play with fire, became an orphan reduced to begging for food. As a lesson for the young audience, the book relates quite carefully the mistakes of little Jane and the remedies that might have spared her house and family, if only she knew better.^[*Little Jane, or The Consequences of Playing with Fire* (Baltimore: sold by H. Vicary, no. 40, Pratt-St, 1824).]  It was a lesson that still bore repeating in mid-nineteenth century America. Mother’s Catechism admonished against using papers to play with fire to avoid “dreadful suffering, if not death” that could come as a consequence.^[  R. Phillips, *The Parent’s Assistant, or Mother’s Catechism; Containing Common Things, Necessary to Be Known by Children at an Early Age*, New edition (New Haven: Published by S. Babcock, 1844).] 

A third significant issue was the ability or inability of the fireplace to warm the rooms it was intended to heat. This heating capacity shaped how inhabitants used the rooms they lived in. As Richard Bushman noted in *Refinement of America*, the seventeenth-century homes built on the model of the traditional hall, pulled people toward the fire for warmth, as much of the heat of these massive fireplaces was drawn up the chimney. Internal temperatures were so cold that liquids froze, even a short distance from the fire. People’s changing expectations of comfort, which Bushman also described, changed their perceptions of whether a room was adequately heated or not.^[Bushman, *The Reﬁnement of America: Persons, Houses, Cities*, 122.]  

One further issue that people living with fireplaces had to content with was “that great domestic nuisance, a smoky chimney.”^[“Art. 23. An Essay on the Construction and Building of Chimneys; Including an Enquiry into the Common Causes of Their Smoking, &c,” *Monthly Review* 62 (May 1780): 408–408.]  It was a perennial concern, with various works dedicated to providing methods to properly build fireplaces and chimneys, as well as works dedicated to diagnosing where the problem lay to cure an already built fireplace or chimney that smoked. A writer in 1833, attributed the suffering from smoky chimneys to “the limited diffusion of science,” and proceeded to describe rules for fixing one cause that a scientist and inventor had proposed more than 30 years earlier.^[“Smoky Chimneys: Economizing Fuel and Lighting Private Dwellings.,” *New-York Farmer & American Gardener’s Magazine* 6, no. 12 (December 1833): 365–67; the writer was referring to the works of Benjamin Thompson, Count Rumford: Benjamin Graf von Rumford, *Collected Works of Count Rumford*, ed. Sanborn Conner Brown, vol. III, 5 vols. (Cambridge, Mass: Belknap Press of Harvard University Press, 1968).]  Other writers praised those who carried out experiments and shared their results for curing the smoky chimney. The question of how to address a smoky chimney effectively was on par with saving fuel in terms of frequency of appearance in the literature of the time.^[See for example, “Smoky Chimneys.,” *Parley’s Magazine: With Fifty Engravings* 1, no. 25 (February 15, 1834): 87; “Smoky Chimneys,” *Cultivator* (Albany) 3, no. 5 (July 1836): 86–87; “Smoky Chimneys and Fire-Places,” *New England Farmer* (1848-1871) 3, no. 6 (March 15, 1851): 102; “Improvement in Fire Places.,” *New England Farmer* 3, no. 17 (November 19, 1824): 134.] 

Given all of these issues, one might expect that fireplaces would disappear from American homes as soon as they were no longer necessary. Yet, that did not happen. A study of Fairfax County, VA, shows us that there is a shift in the number of fireplaces in a home before and after the widespread adoption of central heating, but the majority of homes retained a fireplace.

##Fireplaces in Fairfax County Homes

In order to examine fireplaces in Fairfax County, VA, this study uses real estate assessment records from the county Department of Tax Administration (DTA). I requested access to the records, persuant to the Virginia Freedom of Information Act, and made arrangements with the DTA to download the records through FTP. Maintained by the county as a Microsoft Access database, I exported the data as a fixed-width file, which required significant cleanup in order to work successfully in the R environment. Some of the cleanup was easy to do programmatically, although some of the cleanup had to be done manually, in order to preserve the spacing needed to read the data correctly in the fixed-width file. Inconsistencies such as the # symbol in some records but not others, interchangeable use of tabs and spaces, and other data anomolies were corrected directly in a text editor. Chunking the data into managable file sizes and converting the data into a comma separated value format were done programmatically.

The first tests for converting the data used a short file in order to see what issues came up, then used progressively larger files to find other issues.

There were two separate phases: getting the layout file, which describes the fields and their length, into usable shape, and then using that information to parse the data file into fields. Not all entries in the layout file included information about the length of the field, so the first step was to filter out the layout entries that do not have a value in the LENGTH field.

```{r}
layout <- read.csv("PDLayout.csv", stringsAsFactors = FALSE)
library(dplyr)
layout <- layout %>%
filter(!is.na (LENGTH))

```

Then, read the fixed width format data into a file.
```{r}
PDData <- read.fwf("firstrowsB.txt", widths = layout$LENGTH, col.names = layout$COLUMN.NAME)
head(PDData)
```
